Toward Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-12 paper.tex
Yans updates

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
%% The ARG is mostly tree-like
Compared to organisms like humans which recombine every generation,
in SARS-CoV2 recombination is rare, with recombination nodes accounting
for $<$0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be represented by a
single phylogenetic tree, particularly when looking at
a subset of sampled genomes e.g. restricted by time or limited to close
relatives. We expect the ARG to be particularly tree-like
early in the pandemic, when co-infection was less likely.
\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
\caption{\label{fig:cophylogeny}
Tanglegram comparing the leftmost local tree in the Wide ARG
(sampled to mid-2021)
and an ``all-time'' global NextStrain tree (sampled to January 21, 2023). The
trees are pruned down to the samples present in both the trees. Light grey
lines match the corresponding samples between the two trees; black circles
indicate identical sample partitions between the two trees. Terminal branches
are colour-coded according to the Pango lineage status assigned to the tip
samples; no Pango recombinants are shown, as none are shared between the trees.
The tanglegram was generated using the Neighbor-Net algorithm
\citep{Scornavacca2011-mg} implemented in Dendroscope version 3.8.5
\citep{Huson2012-ys}. See Figure~\ref{fig:cophylogeny_long} for the equivalent
cophylogeny for the Long ARG.}
\end{figure}

@@

Our \texttt{sc2ts} backbone phylogenies do contain small amounts of recombination:
there are 7 recombination nodes in the backbone of the Wide ARG and 8 in the backbone
of the Long ARG. However, the backbones remain extremely tree-like, as these
recombination events involve minor, local topological rearrangements.
All involve recombination between closely related lineages (e.g.\ sibling lineages)
and a number even represent recombination within a single lineage, in which the parent
lineages above a recombination event coalesce with each other before coalescing
with any other lineages in the phylogeny. These could either represent true
recombination events between closely related strains, which have previously gone unnoticed,
or could be artefacts of the current \texttt{sc2ts} method and its parameterization.

Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
a GISAID global ``all-time'' tree from NextStrain (downloaded on January 21, 2023).
We illustrate the backbone phylogeny by visualising
a single tree in the middle of the viral genome
(covering the interval 8393--21618), since other regions of the 
genome show almost identical topologies.
It is clear that the backbone topology of the \texttt{sc2ts} tree shows very close
agreement with the NextStrain tree. The sample genomes cluster by their
assigned Pango lineage status, and many variants and their descendents form
identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
clades, labelled).
Supplementary Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG, with similar results.

@@

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
\caption{\label{fig:recomb_mrcas}
Date of common ancestry between the parents either side of recombination
breakpoints, as a function of degree of parental divergence. MRCAs of parents
inferred to have produced a Pango designated recombinant (XA, XB, etc.) are
identified in orange (A) Data after removing singleton recombinants (XXX parent
pairs). Many recombinations involve parental strains that descend from the same
MRCA: dates for the four most frequently identified MRCA nodes are indicated by
horizontal dotted lines. These frequent MRCA nodes tend to be associated with
major waves and commonly have many immediate children (i.e. are large
polytomies, likely to indicate undersampled outbreaks) (B) Further filtered by
removing HMM inconsistent breakpoints (YYY parent pairs). Note that
Figure~\ref{fig:recomb_mrcas_voc_breakdown} shows the same data broken down by
parental VoC classification.}
\end{figure}

@@

The histograms in Figure~\ref{fig:recomb_mrcas} show that
majority of newly identified recombinants arise from parents that diverged 10 to 30 weeks previously. However, there is also a peak of recombinants that diverged ~95 weeks prior to the event: these correspond to recombinants whose parental common ancestor traces to early 2020, including both Delta-Delta and Delta-Omicron recombinants (Supplementary Figure~\ref{fig:recomb_mrcas_voc_breakdown}). Note that each point in these plots represents a breakpoint, so that recombination events with more than one breakpoint (e.g. with 3 or more parents, which comprise X \% of the recombination events) are represented by several plotted points.

As a check on the reliability of these identified recombination events,
Figure~\ref{fig:recomb_mrcas}B  plots only those breakpoints which are
HMM-consistent. This reduces both the number of recombination nodes /
breakpoints from 1522 / B to C / D in the wide ARG and from 763 / 806 to 459 in
the long ARG (Supplementary Tables 1 and 2).

@@

\subsubsection{Wide ARG}
[Quick overview of the wide ARG.]
In the wide ARG, 44
samples are designated as XA by both Nextclade and GISAID, 237 are designated
as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
After removing singleton recombinants, XA numbers remain unchanged, but XB is
reduced to 235 (229) and XC to 4 (0). We correctly identify that all samples
designated as XB by any method have one or more recombination nodes in their
ancestry.

The simplest Pango recombinant genealogy is XA: all XA samples trace back to a
unique originating recombination node. This is the product of a recombination
between a B.1.177.18 sample (specifically the strain Wales/ALDP-115BF41/2021,
EPI\_ISL: 1012804) which contributed the majority of the genome from the start
to a maximum position of 22227bp, and an unknown (inserted) node with imputed
Pango lineage B.1.1.7, which contributed the remaining right hand portion. The
recombination node has five immediate children: four sample leaves (EPI\_ISLs:
989697, 1019487, 1104468 and 1122630) and a UPGMA-inserted node which is the
ancestor of all other XA samples in the dataset.

For XB, all samples trace back to a originating node which is the product of a
recombination between a B.1 sample (specifically the strain
England/CAMB-7B47D/2020, EPI\_ISL\_433960, which contributed the majority of
the genome from the start up to a maximum position of 23604bp), and two other
UPGMA nodes with imputed Pango lineages B.1.627 (up to a maximum position of
27389bp) and B.1.36.8 (the remaining fragment of the genome). As described
later, in the long ARG the equivalent recombination node has only 2 parents,
with no involvement of  B.1.36.8, and it is possible that the third parent in
the wide ARG is artifactual. Additional complications arise because additional
non-X-designated samples such as B.1.634 also descend from this recombination
(also true in the long ARG, see below), and there are also a small number of
further recombination nodes nested within the XB ``clade''. However, all but
one of these nested recombinations are ancestral to only a negligible fractions
of the designated XB samples. The exception accounts for about 17\% of the
designated XB nodes, and involves a recombination between descendants of the
originating XB recombination. More specifically strain
USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombination
between an XB grandparent and its XB grandchild. This could well be
artifactual.

>>>>

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
%% The ARG is mostly tree-like
Compared to organisms like humans which recombine every generation,
in SARS-CoV2 recombination is rare, with recombination nodes accounting
for $<$0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be represented by a
single phylogenetic tree, particularly when looking at
a subset of sampled genomes e.g. restricted by time or limited to close
relatives. We expect the ARG to be particularly tree-like
early in the pandemic, when co-infection was less likely.

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
\caption{\label{fig:cophylogeny}
Tanglegram comparing a local tree from the Wide ARG
(sampled to mid-2021) and an ``all-time'' global NextStrain tree
(downloaded on January 21, 2023).
Phylogenies are pruned down to those samples present as tips in both datasets. Light grey
lines match the corresponding samples between the two trees; black circles
indicate identical sample partitions between the two trees. Terminal branches
are colour-coded according to the Pango lineage status assigned to the tip
samples; no Pango recombinants are shown, as none are shared between the trees.
The tanglegram was generated using the Neighbor-Net algorithm
\citep{Scornavacca2011-mg} implemented in Dendroscope version 3.8.5
\citep{Huson2012-ys}. See Figure~\ref{fig:cophylogeny_long} for the equivalent
cophylogeny for the Long ARG.}
\end{figure}

@@

Our \texttt{sc2ts} backbone phylogenies do contain small amounts of recombination:
there are 7 recombination nodes in the backbone of the Wide ARG and 8 in the backbone
of the Long ARG. However, the backbones remain extremely tree-like, as these
recombination events involve minor, local topological rearrangements. In particular,
all involve recombination between closely related lineages (e.g.\ sibling lineages)
and a number even represent recombination within a single lineage, in which the parent
lineages above a recombination event coalesce with each other before coalescing
with any other lineages in the phylogeny. These could either represent true
recombination events between closely related strains, which have previously gone unnoticed,
or could be artifacts of the current \texttt{sc2ts} method and its parameterization.

Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
a GISAID global ``all-time'' tree from NextStrain
% We give the date below in the caption - we could probably remove if either from there
% or from here (I suggest from here)
(downloaded on January 21, 2023).
We illustrate the backbone phylogeny by visualizing
a single tree in the middle of the viral genome
%% (covering the interval 8393--21618),  % removed as interval now in the figure title
since other regions of the 
genome show almost identical topologies.
It is clear that the backbone topology of the \texttt{sc2ts} tree shows very close
agreement with the NextStrain tree. The sample genomes cluster by their
assigned Pango lineage status, and many variants and their descendants form
identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
clades, labelled).
Supplementary Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG, with similar results.

@@

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
\caption{\label{fig:recomb_mrcas}
Date of common ancestry between the parents on either side of recombination
breakpoints, as a function of parental divergence. MRCAs of parents
inferred to have caused a Pango designated recombinant (XA, XB, etc.)
identified in orange (A) 851 parent pairs (singleton recombinants removed).
Parental strains often descend from the same MRCA; each horizontal dotted line
indicates a frequently found MRCA: these tend to be associated with major waves
and often have many immediate children (i.e. are large
polytomies, likely to indicate undersampled outbreaks) (B) Further filtered
to 358 parent pairs by removing HMM inconsistent breakpoints. For equivalent
plots broken down by parental VoC classification, see
Figure~\ref{fig:recomb_mrcas_voc_breakdown}
}
\end{figure}

@@

The histograms in Figure~\ref{fig:recomb_mrcas} show that
majority of newly identified recombinants arise from parents that diverged 10 to 30 weeks previously. 
However, there is also a peak of recombinants that diverged ${\sim}95$ weeks prior to the event:
these correspond to recombinants whose parental common ancestor traces to early 2020,
including both Delta-Delta and Delta-Omicron recombinants
(Supplementary Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
Note that each plotted point represents a breakpoint, so that a recombination
event with more than one breakpoint (e.g. with 3 or more parents, comprising
${\sim}10 \%$ of the recombination events in Figure ~\ref{fig:recomb_mrcas}A and
${\sim}3 \%$ in Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.

As a check on the reliability of these identified recombinations,
Figure~\ref{fig:recomb_mrcas}B  plots only those breakpoints which are
HMM-consistent. This reduces both the number of recombination nodes \&
breakpoints in the long ARG from 763 \& 851 to 347 \& 358 (
and from from 1522 \&  to C / D in the wide ARG, see Supplementary Tables 1 and 2).

@@

\subsubsection{Wide ARG}
%% Quick overview of the wide ARG.
In the wide ARG, 44
samples are designated as XA by both Nextclade and GISAID, 237 are designated
as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
After removing singleton recombinants, XA numbers remain unchanged, but XB is
reduced to 235 (229) and XC to 4 (0). We correctly identify that all samples
designated as XB by any method have one or more recombination nodes in their
ancestry.

%% Textual description of the 3 recombinant lineages present in the wide ARG
The simplest Pango recombinant genealogy is XA: all XA samples trace back to a
unique originating recombination node. This is the product of a recombination
between a B.1.177.18 sample (specifically the strain Wales/ALDP-115BF41/2021,
EPI\_ISL: 1012804) which contributed the majority of the genome from the start
to a maximum position of 22227bp, and an unknown (inserted) node with imputed
Pango lineage B.1.1.7, which contributed the remaining right hand portion. The
recombination node has five immediate children: four sample leaves (EPI\_ISLs:
989697, 1019487, 1104468 and 1122630) and a UPGMA-inserted node which is the
ancestor of all other XA samples in the dataset.

For XB, all samples trace back to a originating node which is the product of a
recombination between a B.1 sample (specifically the strain
England/CAMB-7B47D/2020, EPI\_ISL\_433960, which contributed the majority of
the genome from the start up to a maximum position of 23604bp), and two other
UPGMA nodes with imputed Pango lineages B.1.627 (up to a maximum position of
27389bp) and B.1.36.8 (the remaining fragment of the genome). As described
later, in the long ARG the equivalent recombination node has only 2 parents,
with no involvement of  B.1.36.8, and it is possible that the third parent in
the wide ARG is artefactual. Additional complications arise because additional
non-X-designated samples such as B.1.634 also descend from this recombination
(also true in the long ARG, see below), and there are also a small number of
further recombination nodes nested within the XB ``clade''. However, all but
one of these nested recombinations are ancestral to negligible fractions
of the designated XB samples. The exception accounts for about 17\% of the
designated XB nodes, and involves a recombination between descendants of the
originating XB recombination. More specifically strain
USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombination
between an XB grandparent and its XB grandchild. This could well be
artefactual.
